---
title: "R Notebook"
output: html_notebook
---

```{r}
library(tidyverse)
library(plotly)
```

Read in the gapminder_clean.csv data as a tibble using read_csv.

```{r}
gapminder_clean <- read_csv("gapminder_clean.csv")
```
Filter the data to include only rows where Year is 1962 and then make a scatter plot comparing 'CO2 emissions (metric tons per capita)' and gdpPercap for the filtered data.
```{r}
gapminder_1962 <- filter(gapminder_clean, Year==1962)

ggplot(gapminder_1962, aes(x=`CO2 emissions (metric tons per capita)`, y=gdpPercap)) + geom_point()
```

On the filtered data, calculate the correlation of 'CO2 emissions (metric tons per capita)' and gdpPercap. What is the correlation and associated p value?
```{r}
cor.test(gapminder_1962$`CO2 emissions (metric tons per capita)`, gapminder_1962$gdpPercap)
```

On the unfiltered data, answer "In what year is the correlation between 'CO2 emissions (metric tons per capita)' and gdpPercap the strongest?" Filter the dataset to that year for the next step...
```{r}
max_year_cor <- gapminder_clean %>% group_by(Year)%>% 
  summarize(year_cor= cor(`CO2 emissions (metric tons per capita)`, gdpPercap,  use="complete.obs"))  %>% top_n(n=1, wt=year_cor) %>% print()

gapminder_1967 <- filter(gapminder_clean, Year==1967)
```

Using plotly, create an interactive scatter plot comparing 'CO2 emissions (metric tons per capita)' and gdpPercap, where the point size is determined by pop (population) and the color is determined by the continent. You can easily convert any ggplot plot to a plotly plot using the ggplotly() command.
```{r}
p <- ggplot(gapminder_1967, aes(x=`CO2 emissions (metric tons per capita)`, y=gdpPercap, size=pop, color=continent)) + geom_point() 

ggplotly(p)
```

What is the relationship between continent and 'Energy use (kg of oil equivalent per capita)'? (stats test needed)
```{r}

```

Is there a significant difference between Europe and Asia with respect to 'Imports of goods and services (% of GDP)' in the years after 1990? (stats test needed)
```{r}

```

What is the country (or countries) that has the highest 'Population density (people per sq. km of land area)' across all years? (i.e., which country has the highest average ranking in this category across each time point in the dataset?)
```{r}
count_high_pop_dens <- gapminder_clean %>% group_by(`Country Name`) %>% summarize(av_pop_dens = mean(`Population density (people per sq. km of land area)`, na.rm = T)) %>% top_n(n=3, wt=av_pop_dens) %>% print()
```

What country (or countries) has shown the greatest increase in 'Life expectancy at birth, total (years)' since 1962?
```{r}

```

